---
title: "Variables and assignments in R"
author: "Greg Tucker-Kellogg"
date: "January 14, 2019"
output: html_document
teaching: 5
exercises: 0
questions:
- What are variables and why use them?
- How do we assign values to variables?
objectives:
- Understand what variables are and how values are assigned to variables.
- Differentiate global and local variables, and understand how to assign a variable within a function as a global variable. 
- How do we read data into R, or export data from R?
keypoints:
- Variables are used to store values in a script.
- Local variables can only be referred to within the environment it was found in, while
  global environments can be referred to anywhere.
- Variable *values* can be updated by referencing itself and then assigning the new value to itself.
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
source("../bin/chunk-options.R")
```

## R performs calculations

At the R console interface, computed values are returned from every calculation. In this way, R functions like a calculator

```{r}
1 + 1 # Add two numbers
1:30 # generate integers over a range of values
r <- 2 
pi*r^2 # area of a circle with radius 2---

```

## R data are all vectors

A *vector* is a basic data type in R, and contains elements of the same type (e.g., numbers, strings, etc). In each example above, the value returned was shown with an *index* starting with the number `1`. This is because all data, even simple values like numbers, are stored in *vectors*. If we ask for one value, we get a vector of length 1. If we ask for 30 values, we get a vector of length 30.

This is different from many other languages which make a sharp distinction between a simple piece of data and a collection of such pieces. In R, a single number is actually a numeric vector, but of length 1.

> ## R vectors start at 1, not 0
> 
> In R, vectors are indexed beginning with 1. This is different from languages like Java, C, and
> Python, all of which use zero-based indexing. In R, to refer to the first element of a vector `x`,
> use `x[1]`
>
{: .warning}


## Variables: What and why

R uses objects called *variable* to store values. In the third example above, the variable `r` was set to a value of `2` using the assignment operator `<-`. Variables provide a way for a programmer to think in terms of *abstractions* rather than the values that exist at the moment. Thinking about a coding problem in terms of abstractions is a key component of computational thinking.

## Accessing a vector 
